Just Take the Bus: Analysis of Toronto Bike Theft from 2014-2023

### 2024-11-12 Paper/Article.Rmd
Clean up and add figures

---
title: "Just Take the Bus: Analysis of Toronto Bike Theft from 2022-2024"
subtitle: tbd
author:
- Rayan Awad Alim
- Samreen Gill
thanks: https://github.com/RayanAlim/Toronto_Bike_Thefts_Data_Analysis
date: "November 12, 2024"
abstract: This analysis examines Bike theft data obtained from the Tronrot Police Data Portal.
output: pdf_document
toc: true
number-sections: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup}
#| include: false
#| warning: false
#| message: false
# Workspace Set up

knitr::opts_chunk$set(echo=FALSE, include=TRUE, eval=TRUE)
required_packages <- c(
  "lubridate", "sf", "tidyverse", "ggplot2", "maps", "dplyr", "sf", "mapdata", "ggspatial", "opendatatoronto", "here"
) 
for (p in required_packages) { # install if missing
  if (!require(p, character.only = TRUE)) {
    install.packages(p, character.only = TRUE)
  }
}

library(ggplot2)
library(dplyr)
library(tidyverse)
library(sf)
library(mapdata)
library(ggspatial)
library(opendatatoronto)
library(here)
```

```{r}
#| include: false
#| warning: false
#| message: false
# you are welcome to use either the CSV file or the GeoJSON file
# GeoJSON format maybe useful if you would like to combine
# an external geospatial dataset

bike_thefts <- read_csv(here("Data", "Bike_thefts_raw.csv")) |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    )
  )
bike_thefts_sf <- read_sf(here("Data", "Bicycle_Thefts_Open_Data.geojson")) |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    )
  )
```

```{r}
#| include: false
#| warning: false
#| message: false
#glimpse(bike_thefts)
```

@@

How much do they contribute to total theft from 2022 to 2024 ? \
(Comparing with premise type)An in-depth analysis of premise type
unfolds that bicycles last parked outside building premises, whether
residential or commercial are particularly vulnerable to theft. This
trend is consistently observed in top 10 neighborhoods for bike theft.
These locations, especially because of high foot or bike traffic,
provide easy access and quick escape routes for criminals.

\

**When is your bike most likely to be stolen(day of the week, month) ?**

\

@@

```{r}
#BY MONTH
bike_thefts %>%
  count(OCC_MONTH) %>%
  ggplot(aes(x = reorder(OCC_MONTH, n), y = n, fill = OCC_MONTH)) +
  geom_bar(stat = "identity") +
  labs(title = "Bike Thefts by Month of the Year",
       x = "Month", y = "Number of Thefts") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

```{r}
# BY DAY OF WEEK
bike_thefts %>%
  count(OCC_DOW) %>%
  ggplot(aes(x = OCC_DOW, y = n, fill = OCC_DOW)) +
  geom_bar(stat = "identity") +
  labs(title = "Bike Thefts by Day of the Week",
       x = "Day of the Week", y = "Number of Thefts") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3")
```

```{r}
# heaatmap of thefts by day of week and hour

ggplot(bike_thefts, aes(x = OCC_DOW, y = OCC_HOUR)) +
  geom_bin2d(bins = 24) +
  scale_fill_gradient(low = "lightblue", high = "red") +
  labs(title = "Heatmap of Bike Thefts by Day and Hour", 
       x = "Day of Week", 
       y = "Hour of Day") +
  theme_minimal()
```

@@

```{r}
#TOP TEN 
bike_thefts %>%
  count(NEIGHBOURHOOD_158) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(x = reorder(NEIGHBOURHOOD_158, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Neighborhoods for Bike Thefts",
       x = "Neighborhood", y = "Number of Thefts") +
  theme_minimal()
```

```{r}
#| include: false
#| warning: false
#| message: false
#MAP TBD
#library(opendatatoronto)
```

@@

Here we com[are] the frequency of thefts by bike type and cost
distribution:

```{r}
# Barplot for bike types
ggplot(bike_thefts, aes(x = BIKE_TYPE)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Number of Bike Thefts by Bike Type", x = "Bike Type", y = "Count") +
  theme_minimal()
```

```{r}
# Box plot for bike cost distribution by status
ggplot(bike_thefts %>% filter(!is.na(BIKE_COST)), aes(x = STATUS, y = BIKE_COST)) +
  geom_boxplot() +
  labs(title = "Bike Cost Distribution (Recovered vs Stolen)", x = "Status", y = "Cost ($)") +
  theme_minimal()
```

@@

```{r}
#| include: FALSE
#| warning: false
#| message: false
# calcuate reporting delay
bike_thefts <- bike_thefts %>%
  mutate(reporting_delay = as.numeric(difftime(ymd_hms(REPORT_DATE), ymd_hms(OCC_DATE), units = "days")))

ggplot(bike_thefts, aes(x = reporting_delay)) +
  geom_histogram(binwidth = 1, fill = "purple", color = "black") +
  labs(title = "Reporting Delays for Bike Thefts", x = "Days Delay", y = "Count") +
  theme_minimal()
```

@@

```{r}
# calculating recovery rate by neighborhood
recovery_by_neighborhood <- bike_thefts %>%
  group_by(NEIGHBOURHOOD_158) %>%
  summarize(
    total_thefts = n(),
    recovered = sum(STATUS == "RECOVERED"),
    recovery_rate = recovered / total_thefts
  ) %>%
  slice_max(recovery_rate, n = 10) # Get top 10 neighborhoods by recovery rate

#  recovery rate by neighborhood
ggplot(recovery_by_neighborhood, aes(x = reorder(NEIGHBOURHOOD_158, recovery_rate), y = recovery_rate)) +
  geom_bar(stat = "identity", fill = "forestgreen") +
  coord_flip() +
  labs(title = "Top 10 Neighborhoods by Recovery Rate", x = "Neighborhood", y = "Recovery Rate") +
  theme_minimal()
```

```{r}
# map

# Count and filter top 10 neighborhoods with the highest thefts
top_neighborhoods <- bike_thefts %>%
  count(NEIGHBOURHOOD_158) %>%
  slice_max(n, n = 10)

# Treemap plot
treemap::treemap(
  top_neighborhoods,
  index = "NEIGHBOURHOOD_158",  # Using neighborhood names as the index
  vSize = "n",                  # Size by count of thefts
  vColor = "n",                 # Color by count of thefts
  palette = c("#2b5c8a", "goldenrod1", "#9e3d22"),
  type = "manual",
  align.labels = list(c("right", "top"), c("left", "bottom")),
  overlap.labels = 0.2,
  title = "Number of Bicycle Thefts by Neighborhood",
  fontsize.title = 18,
  title.legend = "Frequency",
  fontsize.labels = 10, 
)
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

>>>>

---
title: "Just Take the Bus: Analysis of Toronto Bike Theft from 2014-2023"
subtitle: "Examining Patterns, Locations, and Factors of Bicycle Theft in Toronto"
author:
- Rayan Awad Alim
- Samreen Gill
thanks: https://github.com/RayanAlim/Toronto_Bike_Thefts_Data_Analysis
date: "November 12, 2024"
abstract: This analysis examines Bike theft data obtained from the Tronrot Police Data Portal.
output: pdf_document
toc: true
number-sections: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup}
#| include: false
#| warning: false
#| message: false

# Workspace Set up

knitr::opts_chunk$set(echo=FALSE, include=TRUE, eval=TRUE)
required_packages <- c(
  "lubridate",
  "sf",
  "tidyverse",
  "ggplot2",
  "maps",
  "dplyr",
  "mapdata",
  "ggspatial",
  "opendatatoronto",
  "here",
  "stringr",
  "RColorBrewer",
  "Scales"
) 
for (p in required_packages) { # install if missing
  if (!require(p, character.only = TRUE)) {
    install.packages(p, character.only = TRUE)
  }
}

library(lubridate)
library(sf)
library(tidyverse)
library(ggplot2)
library(maps)
library(dplyr)
library(mapdata)
library(ggspatial)
library(opendatatoronto)
library(here)
library(stringr)
library(RColorBrewer)
library(scales)
```

```{r}
#| include: false
#| warning: false
#| message: false

# Load and process data from CSV and GeoJSON files
bike_thefts <- read_csv(here("Data", "Bike_thefts_raw.csv")) |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    NEIGHBOURHOOD_Clean = str_remove(NEIGHBOURHOOD_158, "\\s*\\(\\d+\\)$")
  )

bike_thefts_sf <- read_sf(here("Data", "Bicycle_Thefts_Open_Data.geojson")) |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    NEIGHBOURHOOD_Clean = str_remove(NEIGHBOURHOOD_158, "\\s*\\(\\d+\\)$")
  )

toronto_shape <- read_sf(here("Data", "Neighbourhoods - 4326.shp")) 
```

```{r}
# glimpse(toronto_shape)
# glimpse(bike_thefts)
```

# Introduction

@@

How much do they contribute to total theft from 2022 to 2024 ? (Comparing with premise type)An in-depth analysis of premise type unfolds that bicycles last parked outside building premises, whether residential or commercial are particularly vulnerable to theft. This trend is consistently observed in top 10 neighborhoods for bike theft. These locations, especially because of high foot or bike traffic, provide easy access and quick escape routes for criminals.

```{r}
# Samreen 
# bar charts of general statistics:
# # of bikes stole in 2021
# # of bikes stole in 2022
# # of bikes stole in 2023
# Total Loss of x $ M 
```

```{r}
# colnames(toronto_shape)
# colnames(thefts_by_neighborhood)
```

```{r}
thefts_by_neighborhood <- bike_thefts %>%
  count(NEIGHBOURHOOD_Clean) %>%
  rename(thefts_count = n)

toronto_shape <- toronto_shape %>%
  rename(NEIGHBOURHOOD_Clean = AREA_NA7)

toronto_thefts_map <- toronto_shape %>%
  left_join(thefts_by_neighborhood, by = "NEIGHBOURHOOD_Clean")

ggplot(toronto_thefts_map) +
  geom_sf(aes(fill = thefts_count), color = "grey", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", name = "Number of Thefts") +
  labs(title = "Number of Bike Thefts per Neighborhood in Toronto") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))
```
premise type:

```{r}
bike_thefts %>%
  count(PREMISES_TYPE) %>%
  ggplot(aes(x = "", y = n, fill = PREMISES_TYPE)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(title = "Bike Theft by Location Type",
       fill = "Location Type") +
  theme_void()
```

## When is your bike most likely to be stolen(day of the week, month) ?

@@

```{r}
# Bike Thefts by Month
bike_thefts <- bike_thefts %>%
  mutate(OCC_MONTH = factor(OCC_MONTH, levels = month.name))

bike_thefts %>%
  count(OCC_MONTH) %>%
  ggplot(aes(x = OCC_MONTH, y = n, fill = OCC_MONTH)) +
  geom_bar(stat = "identity") +
  labs(title = "Bike Thefts by Month of the Year",
       x = "Month", y = "Number of Thefts", fill = "Months") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

```{r}
bike_thefts <- bike_thefts %>%
  mutate(OCC_DOW = factor(OCC_DOW, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")))

bike_thefts %>%
  count(OCC_DOW) %>%
  ggplot(aes(x = OCC_DOW, y = n, fill = OCC_DOW)) +
  geom_bar(stat = "identity") +
  labs(title = "Bike Theft Patterns by Week Day Slight Uptick on Fridays",
       x = "Day of the Week", 
       y = "Number of Thefts", 
       fill = "Day of Week",
       caption = "Bike thefts in Toronto show a consistent pattern across weekdays, with a small increase on Fridays.") +  
  theme_minimal() +
  scale_fill_brewer(palette = "Set3")

```

```{r}
bike_thefts <- bike_thefts %>%
  mutate(OCC_DOW = factor(OCC_DOW, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")))

ggplot(bike_thefts, aes(x = OCC_DOW, y = OCC_HOUR)) +
  geom_bin2d(bins = 24) +
  scale_y_reverse() + 
  scale_fill_gradient(low = "lightblue", high = "red", name = "Number of Thefts") +  
  labs(
    title = "Bike Theft Patterns by Day and Hour",
    x = "Day of the Week",
    y = "Hour of Day",
    caption = "TBD",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

@@

```{r}
bike_thefts %>%
  count(NEIGHBOURHOOD_Clean) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(x = reorder(NEIGHBOURHOOD_Clean, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = n), hjust = 1.2, color = "white") +
  coord_flip() +
  labs(title = "Downtown Danger: Top 10 Neighborhoods for Bike Thefts",
       x = "Neighborhood", y = "Number of Thefts") +
  theme_minimal()
```

```{r}
bike_thefts %>%
  count(NEIGHBOURHOOD_Clean) %>%
  arrange(desc(n)) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(x = "", y = n, fill = reorder(NEIGHBOURHOOD_Clean, -n))) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  geom_text(aes(label = paste0("(", round(n / sum(n) * 100, 1), "%)")),
            position = position_stack(vjust = 0.5),
            color = "white", size = 4) +
  labs(title = "Top 10 Neighborhoods for Bike Thefts",
       fill = "Neighborhood") +
  scale_fill_brewer(palette = "Paired") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))

```

@@

```{r}
total_cost_by_neighborhood <- bike_thefts %>%
  group_by(NEIGHBOURHOOD_Clean) %>%
  summarize(total_cost = sum(BIKE_COST, na.rm = TRUE))

toronto_cost_map <- toronto_shape %>%
  left_join(total_cost_by_neighborhood, by = c("NEIGHBOURHOOD_Clean" = "NEIGHBOURHOOD_Clean"))

ggplot(toronto_cost_map) +
  geom_sf(aes(fill = total_cost), color = "grey", size = 0.1) +
scale_fill_viridis_c(option = "inferno", name = "Total Theft Cost ($)",
                       breaks = seq(0, max(toronto_cost_map$total_cost, na.rm = TRUE), length.out = 10),
                       labels = scales::comma_format()) + 
  labs(title = "Total Cost of Bike Thefts per Neighborhood in Toronto") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))
```

Here we look at the frequency of thefts by bike type:\n
```{r}
total_cost_by_neighborhood %>%
  arrange(desc(total_cost)) %>%
  slice_max(total_cost, n = 10) %>%  
  ggplot(aes(x = reorder(NEIGHBOURHOOD_Clean, total_cost), y = total_cost)) +
  geom_bar(stat = "identity", fill = "tan4") +
  geom_text(aes(label = scales::comma(total_cost)), hjust = 1.2, color = "white") +
  coord_flip() +
  labs(title = "Top 10 Neighborhoods by Total Theft Cost",
       x = "Neighborhood", y = "Total Theft Cost ($)") +
  theme_minimal()
```

```{r}
bike_type_names <- c(
  "BM" = "BMX", 
  "EL" = "Electric Bike", 
  "FO" = "Folding Bike", 
  "MT" = "Mountain Bike", 
  "OT" = "Other Bike", 
  "RC" = "Road Bike", 
  "RE" = "Hybrid Bike",
  "RG" = "??", 
  "SC" = "??", 
  "TA" = "??", 
  "TO" = "??", 
  "TR" = "??", 
  "UN" = "Unspecified"
)

custom_colors <- c("royalblue", "maroon3", "springgreen4", "salmon",
                   "purple4", "wheat2", "maroon", "seagreen1", "khaki4",
                   "lightseagreen", "#9b59b6", "chocolate1", "red3")

ggplot(bike_thefts, aes(x = BIKE_TYPE, fill = BIKE_TYPE)) +
  geom_bar() +
  labs(
    title = "Number of Bike Thefts by Bike Type",
    x = "Bike Type",
    y = "Count",
    fill = "Bike Type"  
  ) +
  scale_fill_manual(values = custom_colors, labels = bike_type_names) +  
  theme_minimal() +
  theme(
    legend.position = "right",  
    axis.text.x = element_text(angle = 45, hjust = 1)  
  )
```

@@

```{r}
# Pie Chart of recovery vs stolen

recovery_status <- bike_thefts %>%
  count(STATUS) %>%
  mutate(percentage = n / sum(n) * 100)

ggplot(recovery_status, aes(x = "", y = n, fill = STATUS)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5),
            color = "white") +
  labs(title = "Bike Recovery Status: Recovered vs Stolen",
       fill = "Status") +
  theme_void() +
  scale_fill_brewer(palette = "Dark2")

```
```{r}
# MAP OF RECOVERY RATE 
recovery_by_neighborhood <- bike_thefts %>%
  group_by(NEIGHBOURHOOD_Clean) %>%
  summarize(
    total_thefts = n(),
    recovered = sum(STATUS == "RECOVERED", na.rm = TRUE),
    recovery_rate = recovered / total_thefts
  )

if (!"sf" %in% class(recovery_by_neighborhood)) {
  toronto_recovery_map <- toronto_shape %>%
    left_join(recovery_by_neighborhood, by = "NEIGHBOURHOOD_Clean")
} else {
  toronto_recovery_map <- st_join(toronto_shape, recovery_by_neighborhood, left = TRUE)
}

ggplot(toronto_recovery_map) +
  geom_sf(aes(fill = recovery_rate), color = "grey") +
  scale_fill_viridis_c(option = "plasma", name = "Recovery Rate") +
  labs(title = "Bike Theft Recovery Rate by Neighborhood in Toronto") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))
```

@@

```{r}
# calculating recovery rate by neighborhood
recovery_by_neighborhood <- bike_thefts %>%
  group_by(NEIGHBOURHOOD_Clean) %>%
  summarize(
    total_thefts = n(),
    recovered = sum(STATUS == "RECOVERED"),
    recovery_rate = recovered / total_thefts
  ) %>%
  slice_max(recovery_rate, n = 10)

ggplot(recovery_by_neighborhood, aes(x = reorder(NEIGHBOURHOOD_Clean, recovery_rate), y = recovery_rate)) +
  geom_bar(stat = "identity", fill = "forestgreen") +
  coord_flip() +
  labs(title = "Top 10 Neighborhoods by Recovery Rate", x = "Neighborhood", y = "Recovery Rate") +
  theme_minimal()
```
